---
title: "Introduction to LaTeX and knitr"
author: "Steve Pederson"
date: "26 August 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LaTeX

### What is LaTex
- LaTeX is a typesetting language
- Brilliant for maths, managing references, tables, figures etc.
- Main file format is `.tex`
- Bibliographies are usually saved as `bibliography.bib`

### Two Key References
- https://www.latex-project.org/
- https://en.wikibooks.org/wiki/LaTeX

## LaTeX | Installation

Installation is big and slow

```
sudo apt-get install texlive-full
```


- Ubuntu: http://www.tug.org/texlive/doc/texlive-en/texlive-en.html
- Mac: http://www.tug.org/mactex/  
- Windows: http://miktex.org/howto/install-miktex

## LaTeX

- We can edit TeX in any IDE, e.g. Texmaker, LyX, Atom, RStudio
- It is not a WYSIWYG editor
- Like markdown but fully featured
    - Control line-spacing, figure size, captions, referencing
    - Generate figures using code
    - Change font sizing
    - Auto generate TOC, Table of Figures, Table of Tables
    - Comment out code (text, figures etc.)

## LaTeX | Using knitr

### Create a File
- In R we use the `Sweave` file format: `Filename.Rnw`
- Let's create one: `File > New File > R Sweave`
- Save as `knitr-LaTeX-tutorial.Rnw`

## LaTeX | The blank template
- The `\documentclass{}` field has set document types
    - `article`, `report`, `letter`, `book` 
    - `beamer` (presentations)
- We can specify the font size, paper size, orientation etc

Set the font size and paper size using:
```
\documentclass[11pt,a4paper]{article}
```

## LaTeX | Basic Text Formatting

We enclose a word in the text formatting we require:

```
These words will be in \textit{italics}, \textbf{bold}, 
\texttt{true-type} and \underline{underlined}.
```

Enter this in your file after  `\begin{document}` then hit  
`Compile PDF`

## LaTeX | Compiling

Numerous files were created during this stage

- Errors will be in the log file
- You can usually ignore the rest.

## LaTeX | Basic Text Formatting

New lines are created by finishing a line using either `\newline` or `\\`

```
These words will be in \textit{italics}, \textbf{bold}, \\
\texttt{true-type} and \underline{underlined}.
```

(Recompile)

## LaTeX | Basic Text Formatting

We can increase the gap before the next line simply

```
These words will be in \textit{italics}, \textbf{bold}, \\[5mm]
\texttt{true-type} and \underline{underlined}.
```

(Recompile)

## LaTeX | Basic Text Formatting

We can insert mathematical symbols and Greek letters in a sentence, using subscripts (_) and superscripts (^)

```
We are testing $H_0 : \sigma^2 \geq 1$
```

(Recompile)

This done in text using `\textsubscript{}` and `\textsuperscript{}`.

## LaTeX | Front Matter

In reality, we'll usually need a header or title.

- This usually goes directly after `\begin{document}`

```
\title{My First LaTeX Document}
\author{Steve Pederson}
\date{August 2016}
\maketitle
```

## LaTeX | Front Matter

You can get fancy in the header if you'd like

```
\title{My First LaTeX Document}
\author{Steve Pederson\\
University of Adelaide\\
e: \texttt{email.me@wherever.com}}
\date{August 2016}
\maketitle
```

## LaTeX | Special Characters

Several characters have special meaning:

```
# $ % ^ & _ { } ~ \
```

- We've already seen the \$ symbol to begin and start in-line maths
- The comment character is %
- All can be printed by escaping, except `\` 
    - This is printed using `\textbackslash{}`

## LaTeX | Special Characters

Try entering these lines in your file

```
I'd like to hide the end of this %line
This is 100\%
```

(Recompile)

## LaTeX | Loading Packages

- Extra functionality is added via `packages`

Under `\documentclass{article}` add
```
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
```

Now we can use a huge range of mathematical symbols

## LaTeX | Document Structure


| Command | Level	| Comment |
| ------- | ----- | ------- |
| `\chapter{''chapter''}` | 0 | only books and reports |
| `\section{''section''}` | 1 | not in letters |
| `\subsection{''subsection''}` | 2 | not in letters |
| `\subsubsection{''subsubsection''}` | 3 | not in letters |
| `\paragraph{''paragraph''}` | 4 | not in letters |

## LaTeX | Document Structure

After `\maketitle` add

```
\section{Introduction}\label{sec:Intro}
```

- By adding the label we can refer to this at any point in the text

(Recompile)

## LaTeX | Referencing

Now later on, add the text:

```
As seen in Section \ref{sec:Intro}, we're awesome
```

- We can reference anything we label, e.g. sections, figures, tables, equations
- If we don't want the numbers at the headings, we can use `\section*{Introduction}`
- We won't be able to use the referencing in the text though

## LaTeX |Table of Contents

If we use numbering, we can make a Table of Contents

- After `\maketitle`, add

```
\tableofcontents
```

(Recompile)

Other options are `\listoffigures` and `\listoftables`

## LaTeX | Environments

- LaTeX works using Environments
- What we've done is create a large environment called a `document`
- Environments are enclosed by `\begin{}` and `\end{}`
- For online maths, we used the shortcut `$...$` to create one

## LaTeX | Environments

- Common environments are `itemize` and `enumerate`
- There are actually shortcuts to these in RStudio

`Hit the TeX Format icon and insert a Numbered List`

- Every additional point needs to begin with `\item`

```
\begin{enumerate}
  \item The first
  \item The second
\end{enumerate}
```

## LaTeX | Tables

Tables usually go in a `tabular` environment

```
\begin{tabular}{llr}
Day & Location & Total\\
\hline 
Monday & Adelaide & 5 \\
Friday & Roseworthy & 10 \\
\end{tabular}
```

## LaTeX | Tables

If we want to number this and caption it

```
\begin{table}
  \center
  \caption{When, where \& how many}
  \begin{tabular}{llr}
  Day & Location & Total\\
  \hline 
  Monday & Adelaide & 5 \\
  Friday & Roseworthy & 10 \\
  \end{tabular}
  \label{tab:Days}
\end{table}
```

## LaTeX and knitr

Code chunks work exactly the same as markdown except:

- Chunks start with `<<chunkName, chunkOptions>>=`
- Chunks end with `@`

## Tables in RStudio

```{r, eval = FALSE}
library(xtable)
xtable(Puromycin, 
       caption = "The sample dataset for Puromycin",
       label = "tab:Puro")
```

We need to set the chunk option `results = 'asis'`
- I usually also set `echo = FALSE, include = TRUE`

## Tables in RStudio

For better formatting:

```{r, eval = FALSE}
library(magrittr)
Puromycin %>%
  xtable(caption = "The sample dataset for Puromycin",
         label = "tab:Puro",
         digits = c(0, 2, 0, 0)) %>%
  print(include.rownames = FALSE,
        caption.placement = "top")
```

Note I had to set digits to zero for the row names

- They were only removed during the `print()` function

## Figures in RStudio

Chunks MUST be named to create figure labels

- The following chunk would create a figure with the label `fig:barplot` (No line-breaks)
- We can then refer to this in the body of text as well
- The default unit is inches

```
<<barplot, echo = FALSE, include = TRUE, 
           fig.width=5, fig.height=4, 
           fig.cap = "A barplot">>=
barplot(1:5)
@
```

## Printing R Output Within Text

To print the output of an R object in text:

- Enclose in `\Sexpr{}` (Symbolic Expression), e.g.

```
Data was collected for \Sexpr{nrow(mtcars)} vehicles
```